---
title: "customer_engagement_analysis"
author: "Ife Abe"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cerulean
---

```{r setup, include=FALSE}
# Set up the environment
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(tidyr)

```

## Introduction
This document analyzes customer engagement data for PalletTrader, focusing on **Buyers** and **Sellers**. The goal is to identify patterns in transaction behavior, customer retention, and high-value customers.

## Generate Synthetic Data
```{r generate-data}
# Set seed for reproducibility
set.seed(123)

# Number of participants
n_customers <- 500

# Generate synthetic dataset
customer_data <- data.frame(
  CustomerID = 1:n_customers,
  CustomerType = sample(c("Buyer", "Seller"), n_customers, replace = TRUE), # Buyer or Seller
  TransactionType = sample(c("Purchase", "Sale"), n_customers, replace = TRUE), # Purchase or Sale
  TransactionCount = sample(1:20, n_customers, replace = TRUE), # Number of transactions
  AvgTransactionValue = round(runif(n_customers, 50, 500), 2), # Transaction value in USD
  SatisfactionScore = sample(1:5, n_customers, replace = TRUE) # Satisfaction on a scale of 1 to 5
)

# Add calculated fields
customer_data <- customer_data %>%
  mutate(
    TotalTransactionValue = TransactionCount * AvgTransactionValue, # Total transaction value
    HighValueCustomer = ifelse(TotalTransactionValue > 3000, "Yes", "No"), # High-value threshold
    Retained = sample(c("Yes", "No"), n_customers, replace = TRUE, prob = c(0.8, 0.2)), # Retention flag
    PrimaryRole = ifelse(CustomerType == "Buyer", "Procurement", "Supply") # Primary activity
  )

# View the first few rows of the dataset
head(customer_data)
```

## 1. Overview of Buyers and Sellers
### **Transaction Metrics by Customer Type**
We will now summarize transaction metrics for **Buyers** and **Sellers**. This includes average transaction count, average transaction value, and total transaction value.

```{r transaction-metrics}
# Summarize transaction metrics by Customer Type
transaction_summary <- customer_data %>%
  group_by(CustomerType) %>%
  summarise(
    AvgTransactionCount = mean(TransactionCount),
    AvgTransactionValue = mean(AvgTransactionValue),
    TotalTransactionValue = sum(TotalTransactionValue),
    AvgSatisfactionScore = mean(SatisfactionScore)
  )

transaction_summary
```

### **Visualizing Transaction Metrics**
We will now visualize the average transaction value and total transaction value for **Buyers** and **Sellers**.

```{r visualize-transaction}
# Bar plot for Average Transaction Value by Customer Type
ggplot(transaction_summary, aes(x = CustomerType, y = AvgTransactionValue, fill = CustomerType)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Transaction Value by Customer Type", y = "Average Transaction Value (USD)") +
  theme_minimal()

# Bar plot for Total Transaction Value by Customer Type
ggplot(transaction_summary, aes(x = CustomerType, y = TotalTransactionValue, fill = CustomerType)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Total Transaction Value by Customer Type", y = "Total Transaction Value (USD)") +
  theme_minimal()
```

## 2. High-Value Customers
### **Identifying High-Value Customers**
We define high-value customers as those with a total transaction value above $3000. Let’s filter out and analyze these customers.

```{r high-value-customers}
# High-value customers based on Total Transaction Value
high_value_customers <- customer_data %>%
  filter(HighValueCustomer == "Yes")

# Summary of High-Value Customers
high_value_summary <- high_value_customers %>%
  group_by(CustomerType) %>%
  summarise(
    AvgTotalTransactionValue = mean(TotalTransactionValue),
    AvgSatisfactionScore = mean(SatisfactionScore)
  )

high_value_summary
```

### **Visualizing High-Value Customers**
We will visualize the proportion of high-value customers for **Buyers** and **Sellers**.

```{r visualize-high-value}
# Pie chart for proportion of High-Value Buyers vs. Sellers
high_value_count <- customer_data %>%
  group_by(CustomerType, HighValueCustomer) %>%
  tally() %>%
  filter(HighValueCustomer == "Yes")

ggplot(high_value_count, aes(x = "", y = n, fill = CustomerType)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Proportion of High-Value Customers by Type") +
  theme_void()
```

## 3. Retention Analysis
### **Retention by Customer Type**
Let’s now analyze customer retention for **Buyers** and **Sellers**, comparing retention rates between these groups.

```{r retention-analysis}
# Retention analysis by Customer Type
retention_summary <- customer_data %>%
  group_by(CustomerType, Retained) %>%
  tally() %>%
  pivot_wider(names_from = Retained, values_from = n, values_fill = list(n = 0))

# Calculate retention rates
retention_summary <- retention_summary %>%
  mutate(
    RetentionRate = `Yes` / (`Yes` + `No`)
  )

retention_summary

```

### **Visualizing Retention Rates**
We will visualize the retention rates for **Buyers** and **Sellers**.

```{r visualize-retention}
# Bar plot for Retention Rates by Customer Type
ggplot(retention_summary, aes(x = CustomerType, y = RetentionRate, fill = CustomerType)) +
  geom_bar(stat = "identity") +
  labs(title = "Retention Rate by Customer Type", y = "Retention Rate") +
  theme_minimal()
```
```

---

## Conclusion
The analysis of customer data for PalletTrader shows significant differences in transactional behaviors and retention rates between **Buyers** and **Sellers**. High-value customers can be identified based on their total transaction value, and customer retention is an important factor to explore for future growth.
